First Order Alternation

### 2018-06-08 quantifiers/draft.tex
ok

For two integers $0 \leq i \leq j$, we denote by $[i,j]$ the set
$\set{i,i+1,\ldots,j}$ and by $[i]$ the set $[0,i]$. To simplify the
presentation, we consider two sorts $\Data$ and $\Bool$, where $\Data$
ranges over an infinite domain and $\Bool$ has the usual
interpretation. The $\Data$ sort is equipped with a finite set of
function symbols $f : \Data^{\#(f)} \rightarrow \Data$, where
$\#(f)\geq0$ denotes the number of arguments (arity) of $f$. When
$\#(f)=0$, we say that $f$ is a constant, and denote by $\top$ and
$\bot$ the only constants of sort $\Bool$, representing the true and
false values, respectively. A \emph{predicate} is a function symbol
$p : \Data^{\#(p)} \rightarrow \Bool$, denoting a relation of arity
$\#(p)$ and we write $\preds$ for the set of predicates.

@@

An \emph{interpretation} $\I$ maps each predicate $p$ into a set $p^\I
\subseteq \Data^{\#(p)}$, if $\#(p)>0$, or into an element of $\Data$
if $\#(p)=0$. A \emph{valuation} $\nu$ maps each variable $x$ into an
element of $\Data$. Given a term $t$, we denote by $t^\nu$ the value
obtained by replacing each variable $x$ by the value $\nu(x)$ and
evaluating each function application. For a formula $\phi$, we define
the forcing relation $\I,\nu \models \phi$ recursively on the
structure of $\phi$, as follows:
\[
\begin{array}{rcl}
\I,\nu \models t \teq s & \iff & t^\nu=s^\nu \\
\I,\nu \models t \not\teq s & \iff & t^\nu \neq s^\nu \\
\I,\nu \models \phi_1 \wedge \phi_2 & \iff & \I,\nu \models \phi_i \text{, for all } i=1,2 \\ 
\I,\nu \models \phi_1 \vee \phi_2 & \iff & \I,\nu \models \phi_i \text{, for some } i=1,2 \\
\I,\nu \models \forall x ~.~ \phi_1 & \iff & \I,\nu[x\leftarrow d] \models \phi_1 \text{, for all } d \in \Data \\
\I,\nu \models \exists x ~.~ \phi_1 & \iff & \I,\nu[x\leftarrow d] \models \phi_1 \text{, for some } d \in \Data \\
\I,\nu \models p(t_1,\ldots,t_{\#(p)}) & \iff & \tuple{t_1^\nu,\ldots,t_{\#(p)}^\nu} \in p^\I 
\end{array}
\]
For a formula $\phi$, we define $\sem{\phi}_\nu \isdef \set{\I \mid
  \I,\nu \models \phi}$, where $\nu(x)$ is defined for each $x \in
\fv{}{\phi}$, and drop the subscript for sentences. An element of
$\sem{\phi}$ is called a \emph{model} of $\phi$. A formula is
satisfiable if $\sem{\phi}\neq\emptyset$.

Interpretations are partially ordered by the pointwise subset order,
defined as $\I_1 \subseteq \I_2$ if and only if $p^{\I_1} \subseteq
p^{\I_2}$ for each predicate $p \in \preds$. Given a set $\mathcal{M}$
of interpretations, a minimal element $\I \in \mathcal{M}$ is an
interpretation such that for no other interpretation $\I' \in
\mathcal{M} \setminus \set{\I}$ do we have $\I' \subseteq \I$. For a
formula $\phi$ and a valuation $\nu$, we denote by $\minsem{\phi}_\nu$
and $\minsem{\phi}$ the set of minimal interpretations from
$\sem{\phi}_\nu$ and $\sem{\phi}$, respectively.

@@

Let $\Sigma$ be a finite alphabet $\Sigma$ of \emph{input
  events}. Given a finite set of variables $X$, we denote by $X
\mapsto \Data^\I$ the set of valuations of the variables $X$ and
$\Sigma[X] = \Sigma \times (X \mapsto \Data)$ be the possibly infinite
set of \emph{data symbols}. A \emph{data word} (word) is a finite
sequence $(a_1,\nu_1)(a_2,\nu_2) \ldots (a_n,\nu_n)$ of data symbols,
where $a_1,\ldots,a_n \in \Sigma$ and $\nu_1,\ldots,\nu_n : X
\rightarrow \Data$ are valuations. We denote by $\varepsilon$ the
empty sequence, by $\Sigma^*$ the set of finite sequences of input
events and by $\Sigma[X]^*$ the set of data words over the variables
$X$.

@@

The execution semantics of of automata is given in close analogy with
the case of boolean alternating automata, with transition rules of the
form $q \arrow{a}{} \phi$, where $q$ is a boolean constant and $\phi$
a positive boolean combination of such constants. For instance, $q_0
\arrow{a}{} q_1 \wedge q_2 \vee q_3$ means that the automaton can
choose to transition in either both $q_1$ and $q_2$ or in $q_3$
alone. This intuition leads to saying that the steps of the automaton
are defined by the minimal boolean models of the transition
formulae. In this case, both $\set{q_1 \leftarrow \top, q_2 \leftarrow
  \top, q_3 \leftarrow \bot}$ and $\set{q_1 \leftarrow \bot, q_2
  \leftarrow \bot, q_3 \leftarrow \top}$ are minimal models, but not
$\set{q_1 \leftarrow \top, q_2 \leftarrow \top, q_3 \leftarrow \top}$,
which is a model but is not minimal.

Given a predicate $q \in Q$ and a tuple of data values
$d_1,\ldots,d_{\#(q)}$, the tuple $q(d_1,\ldots,d_{\#(q)})$ is called
a \emph{configuration}\footnote{Note that a configuration is not a
  logical term since data values cannot be written in logic.}. To
formalize the execution semantics of automata, we relate sets of
configurations to models of first-order sentences, as follows. Given a
formula $\phi$ and a valuation of its free variables $\nu$, each
interpretation $\I \in \sem{\phi}_\nu$ corresponds to a set of
configurations $\cube{\I} \isdef \set{q(d_1,\ldots,d_{\#(q)}) \mid
  \tuple{d_1,\ldots,d_{\#(q)}} \in q^\I}$, called a \emph{cube}. For a
set $\mathcal{M}$ of interpretations, we define $\cube{\mathcal{M}}
\isdef \set{\cube{\I} \mid \I \in \mathcal{M}}$. Given a word
$w=(a_1,\nu_1)(a_2,\nu_2) \ldots (a_n,\nu_n) \in \Sigma[X]^*$, an
\emph{execution} of $\A=\tuple{\Sigma,X,Q,\iota,F,\Delta}$ over $w$ is
a (possibly infinite) forest $\F_w = \set{T_1,T_2,\ldots}$, where each
$T_i$ is defined below:
\begin{compactitem}
\item $T_i$ is rooted in a configuration from some cube from
  $\cube{\minsem{\iota}}$ --- because $\iota$ is satisfiable, such cube
  always exists,
%
\item all paths from the root to a leaf of $T_i$ have length $n$, and
%
\item if $q(d_1,\ldots,d_{\#(q)})$ is a node on level $j \in [0,n-1]$
  in $T_i$, then its children form a cube from
  $\cube{\minsem{\psi}_\eta}$, where \(q(y_1,\ldots,y_{\#(q)})
  \arrow{a_{j+1}(X)}{} \psi \in \Delta\) is a transition rule and
  $\eta$ is a valuation that assigns each variable $y_k$ the value
  $d_k$, for $k \in [1,\#(q)]$ and each variable $x \in X$ the value
  $\nu_{j+1}(x)$ --- because $\fv{}{\psi} \subseteq X \cup
  \set{y_1,\ldots,y_{\#(q)}}$, the values assigned by $\eta$ to
  variables outside $X \cup \set{y_1,\ldots,y_{\#(q)}}$ have no impact
  on $\cube{\minsem{\psi}_\eta}$.
\end{compactitem}
A forest $\F_w$ is said to be \emph{accepting} if and only if the
frontier of each tree $T \in \F_w$ is labeled with configurations
$q(d_1,\ldots,d_{\#(q)})$ such that $q \in F$ is a final predicate. In
this case, we say that $\A$ accepts $w$ and denote by $\lang{\A}$ the
set of words accepted by $\A$. In this paper, we answer the following
questions: \begin{compactenum}
\item \emph{closure}: given automata $\A_i =
  \tuple{\Sigma,X,Q_i,\iota_i,F_i,\Delta_i}$, for $i=1,2$, do there
  exist automata $\A_\cap$, $\A_\cup$ and $\overline{\A}_1$ such that
  $L(\A_\cap) = L(\A_1) \cap L(\A_2)$, $L(\A_\cup) = L(\A_1) \cup
  L(\A_2)$ and $L(\overline{\A}_1) = \Sigma[X]^* \setminus L(\A_1)$ ?
%
\item \emph{emptiness}: given an automaton $\A$, is $L(\A) =
  \emptyset$?
\end{compactenum}

@@

>>>>

For two integers $0 \leq i \leq j$, we denote by $[i,j]$ the set
$\set{i,i+1,\ldots,j}$ and by $[i]$ the set $[0,i]$. To simplify the
presentation, we consider two sorts $\Data$ and $\Bool$, where $\Data$
ranges over an infinite domain and $\Bool$ has the usual
interpretation. The $\Data$ sort is equipped with a finite set of
function symbols $f : \Data^{\#(f)} \rightarrow \Data$, where
$\#(f)\geq0$ denotes the number of arguments (arity) of $f$. When
$\#(f)=0$, we say that $f$ is a constant, and denote by $\top$ and
$\bot$ the only constants of sort $\Bool$, representing the true and
false values, respectively. A \emph{predicate} is a function symbol $p
: \Data^{\#(p)} \rightarrow \Bool$, denoting a relation of arity
$\#(p)$ and we write $\preds$ for the set of predicates. In the
following, we shall consider that the interpretation of all function
symbols $f : \Data^{\#(f)} \rightarrow \Data$ that are not predicates
is fixed and relative to the interpretation of the $\Data$ sort.

@@

An \emph{interpretation} $\I$ maps each predicate $p$ into a set $p^\I
\subseteq \Data^{\#(p)}$, if $\#(p)>0$, or into an element of $\Data$
if $\#(p)=0$. A \emph{valuation} $\nu$ maps each variable $x$ into an
element of $\Data$. Given a term $t$, we denote by $t^\nu$ the value
obtained by replacing each variable $x$ by the value $\nu(x)$ and
evaluating each function application. For a formula $\phi$, we define
the forcing relation $\I,\nu \models \phi$ recursively on the
structure of $\phi$, as follows:
\[
\begin{array}{rcl}
\I,\nu \models t \teq s & \iff & t^\nu=s^\nu \\
\I,\nu \models t \not\teq s & \iff & t^\nu \neq s^\nu \\
\I,\nu \models p(t_1,\ldots,t_{\#(p)}) & \iff & \tuple{t_1^\nu,\ldots,t_{\#(p)}^\nu} \in p^\I \\
\I,\nu \models \phi_1 \wedge \phi_2 & \iff & \I,\nu \models \phi_i \text{, for all } i=1,2 \\ 
\I,\nu \models \phi_1 \vee \phi_2 & \iff & \I,\nu \models \phi_i \text{, for some } i=1,2 \\
\I,\nu \models \forall x ~.~ \phi_1 & \iff & \I,\nu[x\leftarrow d] \models \phi_1 \text{, for all } d \in \Data \\
\I,\nu \models \exists x ~.~ \phi_1 & \iff & \I,\nu[x\leftarrow d] \models \phi_1 \text{, for some } d \in \Data 
\end{array}
\]
For a formula $\phi$, we define $\sem{\phi}_\nu \isdef \set{\I \mid
  \I,\nu \models \phi}$, where $\nu(x)$ is defined for each $x \in
\fv{}{\phi}$, and drop the subscript for sentences. A sentence $\phi$
is \emph{satisfiable} if $\sem{\phi}\neq\emptyset$, an element of
$\sem{\phi}$ being called a \emph{model} of $\phi$.

Interpretations are partially ordered by the pointwise subset order,
defined as $\I_1 \subseteq \I_2$ if and only if $p^{\I_1} \subseteq
p^{\I_2}$ for each predicate $p \in \preds$. Given a set $\mathcal{S}$
of interpretations, a minimal element $\I \in \mathcal{S}$ is an
interpretation such that for no other interpretation $\I' \in
\mathcal{S} \setminus \set{\I}$ do we have $\I' \subseteq \I$. For a
formula $\phi$ and a valuation $\nu$, we denote by $\minsem{\phi}_\nu$
and $\minsem{\phi}$ the set of minimal interpretations from
$\sem{\phi}_\nu$ and $\sem{\phi}$, respectively.

@@

Let $\Sigma$ be a finite alphabet $\Sigma$ of \emph{input
  events}. Given a finite set of variables $X$, we denote by $X
\mapsto \Data^\I$ the set of valuations of the variables $X$ and
$\Sigma[X] = \Sigma \times (X \mapsto \Data)$ be the possibly infinite
set of \emph{data symbols}. A \emph{data word} (word) is a finite
sequence $w=(a_1,\nu_1)(a_2,\nu_2) \ldots (a_n,\nu_n)$ of data
symbols, where $a_1,\ldots,a_n \in \Sigma$ and $\nu_1,\ldots,\nu_n : X
\rightarrow \Data$ are valuations. Given a word $w$, we denote by
$\event{w} \isdef a_1\ldots a_n$ its sequence of input events and by
$\data{w}$ the valuation associating each time-stamped variable
$\stamp{x}{i}$ the value $\nu_i(x)$, for all $x \in \vars$ and
$i\in[1,n]$. We denote by $\varepsilon$ the empty sequence, by
$\Sigma^*$ the set of finite sequences of input events and by
$\Sigma[X]^*$ the set of data words over the variables $X$.

@@

The execution semantics of of automata is given in close analogy with
the case of boolean alternating automata, with transition rules of the
form $q \arrow{a}{} \phi$, where $q$ is a boolean constant and $\phi$
a positive boolean combination of such constants. For instance, $q_0
\arrow{a}{} q_1 \wedge q_2 \vee q_3$ means that the automaton can
choose to transition in either both $q_1$ and $q_2$ or in $q_3$
alone. This intuition leads to saying that the steps of the automaton
are defined by the minimal boolean models of the transition
formulae. In this case, both $\set{q_1 \leftarrow \top, q_2 \leftarrow
  \top, q_3 \leftarrow \bot}$ and $\set{q_1 \leftarrow \bot, q_2
  \leftarrow \bot, q_3 \leftarrow \top}$ are minimal models, but not
$\set{q_1 \leftarrow \top, q_2 \leftarrow \top, q_3 \leftarrow \top}$,
which is a model but is not minimal\footnote{To avoid the confusion of
  considering potentially non-minimal models, the original definition
  of alternating finite-state automata \cite{ChandraKozenStockmeyer81}
  uses boolean valuations instead of formulae.}.

Given a predicate $q \in Q$ and a tuple of data values
$d_1,\ldots,d_{\#(q)}$, the tuple $q(d_1,\ldots,d_{\#(q)})$ is called
a \emph{configuration}\footnote{Note that a configuration is not a
  logical term since data values cannot be written in logic.}. To
formalize the execution semantics of automata, we relate sets of
configurations to models of first-order sentences, as follows. Given a
formula $\phi$ and a valuation of its free variables $\nu$, each
interpretation $\I \in \sem{\phi}_\nu$ corresponds to a set of
configurations $\cube{\I} \isdef \set{q(d_1,\ldots,d_{\#(q)}) \mid
  \tuple{d_1,\ldots,d_{\#(q)}} \in q^\I}$, called a \emph{cube}. For a
set $\mathcal{S}$ of interpretations, we define $\cube{\mathcal{S}}
\isdef \set{\cube{\I} \mid \I \in \mathcal{S}}$. Given a word
$w=(a_1,\nu_1)(a_2,\nu_2) \ldots (a_n,\nu_n) \in \Sigma[X]^*$, an
\emph{execution} of $\A=\tuple{\Sigma,X,Q,\iota,F,\Delta}$ over $w$ is
a (possibly infinite) forest $\F_w = \set{T_1,T_2,\ldots}$, where each
tree $T_i$ is defined below:
\begin{compactitem}
\item $T_i$ is rooted in a configuration from some cube from
  $\cube{\minsem{\iota}}$ --- because $\iota$ is satisfiable, such a cube
  always exists,
%
\item all paths from the root to a leaf of $T_i$ have length $n$, and
%
\item if $q(d_1,\ldots,d_{\#(q)})$ is a node on the level $j \in
  [n-1]$ in $T_i$, then its children form a cube from
  $\cube{\minsem{\psi}_\eta}$, where \(q(y_1,\ldots,y_{\#(q)})
  \arrow{a_{j+1}(X)}{} \psi \in \Delta\) is a transition rule and
  $\eta$ is a valuation that assigns each variable $y_k$ the value
  $d_k$, for $k \in [1,\#(q)]$ and each variable $x \in X$ the value
  $\nu_{j+1}(x)$ --- because $\fv{}{\psi} \subseteq X \cup
  \set{y_1,\ldots,y_{\#(q)}}$, the values assigned by $\eta$ to
  variables outside $X \cup \set{y_1,\ldots,y_{\#(q)}}$ have no impact
  on $\cube{\minsem{\psi}_\eta}$.
\end{compactitem}
A forest $\F_w$ is said to be \emph{accepting} if and only if the
frontier of each tree $T \in \F_w$ is labeled with configurations
$q(d_1,\ldots,d_{\#(q)})$ such that $q \in F$ is a final predicate. In
this case, we say that $\A$ accepts $w$ and denote by $\lang{\A}$ the
set of words accepted by $\A$. In this paper, we answer the following
questions: \begin{compactenum}
\item \emph{closure}: given automata $\A_i =
  \tuple{\Sigma,X,Q_i,\iota_i,F_i,\Delta_i}$, for $i=1,2$, do there
  exist automata $\A_\cap$, $\A_\cup$ and $\overline{\A}_1$ such that
  $L(\A_\cap) = L(\A_1) \cap L(\A_2)$, $L(\A_\cup) = L(\A_1) \cup
  L(\A_2)$ and $L(\overline{\A}_1) = \Sigma[X]^* \setminus L(\A_1)$ ?
%
\item \emph{emptiness}: given an automaton $\A$, is $L(\A) =
  \emptyset$?
\end{compactenum}

@@

In the upcoming developments it is more convenient to work with path
formulae defining executions.  Let $\A =
\tuple{\Sigma,X,Q,\iota,F,\Delta}$ be an automaton for the rest of
this session. For any $i \in \nat$, we denote by $\stamp{Q}{i} =
\set{\stamp{q}{i} \mid q \in Q}$ and $\stamp{X}{i} = \set{\stamp{x}{i}
  \mid x \in X}$ the sets of time-stamped predicates and variables,
respectively. For a formula $\psi$ and $i \in \nat$, we define
$\stamp{\psi}{i} \isdef \psi[\stamp{X}{i}/X,\stamp{Q}{i}/Q]$ the
formula in which all input variables and state predicates are replaced
by their time-stamped counterparts.

Given a sequence of input events $\alpha = a_1 \ldots a_n$, the
\emph{path formula} of $\alpha$ in $\A$ is defined as follows: 
\[\pathform{\A}{\alpha} \isdef \stamp{\iota}{0} \wedge 
\bigwedge_{i=1}^n \bigwedge_{q(y_1,\ldots,y_{\#(q)}) \arrow{a_i(X)}{}
  \psi \in \Delta} \forall y_1 \ldots \forall y_{\#(q)} ~.~
\stamp{q}{i-1}(y_1,\ldots,y_{\#(q)}) \Longrightarrow \stamp{\psi}{i}\]
Observe that path formulae are not positive, due to the occurrence of
predicate atoms to the left of the implication as in
$\stamp{q}{i-1}(y_1,\ldots,y_{\#(q)}) \Longrightarrow
\stamp{\psi}{i}$. The intuition is that, whenever the automaton is in
a configuration $q(d_1, \ldots, d_{\#(q)})$ at step $i-1$, it proceeds
at step $i$, according to $\psi$.

To formalize the relation between the low-level configuration-based
execution semantics and the symbolic path formulae, consider a word
$w=(a_1,\nu_1) \ldots (a_n,\nu_n)$. Any execution forest $\F_w$ of
$\A$ over $w$ is associated a unique interpretation defined as
$\I_{[\F_w]}(\stamp{q}{i}) \isdef \set{\tuple{d_1,\ldots,d_{\#(q)}}
  \mid q(d_1,\ldots,d_{\#(q)}) \text{ occurs on level $i$ in $\F_w$}}$
for each $q \in \preds$ and $i \in [n]$.

\begin{lemma}\label{lemma:path-formula}
  For any word $w=(a_1,\nu_1) \ldots (a_n,\nu_n)$, we have: 
  \[\minsem{\pathform{\A}{w_\Sigma}}_{w_\Data} = \set{\I_{[\F_w]} \mid \F_w
    \text{ is an execution of $\A$ over $w$}}\enspace.\]
\end{lemma}

\comment[ri]{I am here}
